Relative parabolic factorizations of Chevalley groups

### 2015-05-06 bass-kolster.tex
Minor changes
Fixed statement of Bass-Kolster; improved treatment of representations

\begin{thm}\label{thm:BassKolster}
Let $\Phi$ be a classical root system of rank $\ell\geqslant2$, let $R$ be arbitary commutative ring and $I$ be an ideal, satisfying one of the following assumptions:
\newline \indent \begin{tabular}{l l l}      
\textbullet & $\Phi = \rA_\ell$, $\ell\geq 2$, & $\sr(I) \leqslant \ell$; \\     
\textbullet & $\Phi = \rB_\ell$, $\rD_\ell$ $\ell\geq 3$,  & $\asr(I) \leqslant \ell-1$; \\ 
\textbullet & $\Phi = \rC_\ell$, $\ell\geq 2$, & $\sr(I) \leqslant 2\ell-1$; \\ 
\end{tabular}

\noindent Then the principal congruence subgroup $\G(\Phi,R,I)$ admits the following relative version of Bass---Kolster decomposition:
\[\G(\Phi,R,I)=\U(\Sigma_1^-\setminus\{-\alpha_\mathrm{max}\},I)\cdot \U(\Sigma_1^+,I)\cdot \Z\cdot \U(\Sigma_1^-,I)\cdot \U(\Sigma_1^+,I)\cdot \G(\Delta_1,R,I), \]
where $\Z=\left\{ z_{-\alpha_\mathrm{max}}(r,1)\mid r\in I \right\}$.
\end{thm}
\begin{proof}
Let $g$ be an element of $\G(\Phi, R, I)$. It is clear that vector $v=g\cdot v^+$ is $I$-unimodular.
In view of relative Chevalley---Matsumoto decomposition it suffices to show that $\Z \cdot \U(\Phi^-, I) \cdot \U(\Phi^+, I) \cdot g$ contains an element whose matrix entry at position $(1,1)$ is equal to $1$.
We will need to analyze several cases.

\textsc{Case $\Phi=\rA_\ell$:}
Denote $v=(1+v_1,v_2,\ldots,v_\ell,v_{\ell+1})^t\in\Ums(\ell+1,R,I)$.
One can add the last entry to the first $\ell$ with some coefficients from $I$ to make them form a unimodular column of length $\ell$.
Then $\ell$ lower transformations (with coefficients from $I$) can make the last entry equal to $v_1$.
Now apply $z_\gamma(-v_1,1)$, $\gamma=-\alpha_{\max}$:
\begin{multline*}
(1+v_1,v_2,\ldots,v_\ell,v_1)^t\xmapsto{x_{-\gamma}(-1)}(1,v_2,\ldots,v_\ell,v_1)^t\xmapsto{x_\gamma(-v_1)}\\\longmapsto(1,v_2,\ldots,v_\ell,0)^t\xmapsto{x_{-\gamma}(1)}(1,v_2,\ldots,v_\ell,0)^t.
\end{multline*}

>>>>

\begin{thm}\label{thm:BassKolster}
Let $\Phi$ be a classical root system of rank $\ell\geqslant2$, let $R$ be arbitary commutative ring and $I$ be an ideal, satisfying one of the following assumptions:
\newline \indent \begin{tabular}{l l l}      
\textbullet & $\Phi = \rA_\ell$, $\ell\geq 2$, & $\sr(I) \leqslant \ell$; \\     
\textbullet & $\Phi = \rC_\ell$, $\ell\geq 2$, & $\sr(I) \leqslant 2\ell-1$; \\ 
\textbullet & $\Phi = \rB_\ell$, $\rD_\ell$ $\ell\geq 3$,  & $\asr(I) \leqslant \ell-1$. \end{tabular}

\noindent Then the principal congruence subgroup $\G(\Phi,R,I)$ admits the following relative version of Bass---Kolster decomposition:
\[\G(\Phi,R,I)=  \G(\Delta_1,R,I)\cdot \U(\Sigma_1^+,I)\cdot \U(\Sigma_1^-\setminus\{-\alpha_\mathrm{max}\},I)\cdot\Z \cdot \U(\Phi^-,I)\cdot \U(\Phi^+,I), \]
where $\Z=\left\{ z_{-\alpha_\mathrm{max}}(r,1)\mid r\in I \right\}$.
\end{thm}
\begin{proof}
Let $g$ be an element of $\G(\Phi, R, I)$. It is clear that vector $v=g\cdot v^+$ is $I$-unimodular.
By relative Chevalley---Matsumoto lemma it suffices to show that the set $\Z \cdot \U(\Phi^-, I) \cdot \U(\Phi^+, I) \cdot g$ contains an element $h$ such that $\mu_1(h\cdot v^+)=1$, $\mu_{\varpi\ssub{1} - \alpha\ssub{max}}(h \cdot v^+)=0$.
We find such $h$ by analyzing several cases.

\textsc{Case $\Phi=\rA_\ell$:}
Set $v=(1+v_1,v_2,\ldots,v_\ell,v_{\ell+1})^t\in\Ums(\ell+1,R,I)$.
One can add the last entry to the first $\ell$ with some coefficients from $I$ to make them form a unimodular column of length $\ell$.
Then $\ell$ lower transformations (with coefficients from $I$) can make the last entry equal to $v_1$.
Now apply $z_\gamma(-v_1,1)$, $\gamma=-\alpha_{\max}$:
\begin{multline*}
(1+v_1,v_2,\ldots,v_\ell,v_1)^t\xmapsto{x_{-\gamma}(-1)}(1,v_2,\ldots,v_\ell,v_1)^t\xmapsto{x_\gamma(-v_1)}\\\longmapsto(1,v_2,\ldots,v_\ell,0)^t\xmapsto{x_{-\gamma}(1)}(1,v_2,\ldots,v_\ell,0)^t.
\end{multline*}

### 2015-05-06 paper.tex
Minor changes
Fixed statement of Bass-Kolster; improved treatment of representations

In this paper we work with irreducible representations of Chevalley groups corresponding to the highest weight
$\varpi_1$ (in the case of a classical $\Phi_l$) or $\varpi_l$ (in the case $\Phi_\ell=\rE_\ell$). 
The former are the natural vector representations of classical groups acting on free modules $V=R^n$ of dimension 
$n=\ell+1, 2\ell+1, 2\ell,2\ell$ for $\Phi=\rA_\ell,\rB_\ell,\rC_\ell,\rD_\ell$.
The latter representations have dimension $n=27, 56, 248$ for $\ell=6,7,8$.

For a representation $\pi$ of $\G(\Phi, R)$ on $V$ we denote by $\Lambda(\pi)$ the set of weights of $\pi$.
It will be convenient to use standard numbering for weights of representation $\pi$ in the case when $\Phi$ is classical and $\pi$ is a vector representation (cp.~\cite[§1B]{St78}):
$$\begin{array}{cll}
  1,2,\ldots, l+1 & \text{in the case} & \Phi =\rA_l, \\
  1,2,\ldots l, 0, -\ell,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_\ell, \\
  1,2,\ldots l, -\ell,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_\ell, \rD_\ell. \\
\end{array}$$
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ and so on.

In what follows we also assume that there is some basis $E=\{e_\lambda\}_{\lambda\in\Lambda(\pi)}$ fixed for every representation $V$.
We denote by $v^+$ the basis vector corresponding to highest weight. 
Using the the above notation for classical $\Phi$ we have $v^+=e_1$.
%TODO: State Matsumoto lemma and Chevalley-Matsumoto lemma

@@

\section{Relative parabolic factorizations}
\subsection{Relative Bass---Kolster decompositions}\label{sec:bass-kolster}
\input{bass-kolster}
\subsection{Relative Dennis-Vaserstein decompositions}\label{sec:dennis-vaserstein}
\input{dennis-vaserstein}

>>>>

In this paper we work with irreducible representations of Chevalley groups corresponding to the highest weight
$\varpi_1$ (in the case of a classical $\Phi_l$) or $\varpi_l$ (in the case $\Phi_\ell=\rE_\ell$). 
The former are the natural vector representations of classical groups acting on free modules $V=R^n$ of dimension 
$n=\ell+1, 2\ell+1, 2\ell,2\ell$ for $\Phi=\rA_\ell,\rB_\ell,\rC_\ell,\rD_\ell$.
The latter representations have dimension $n=27, 56$ for $\ell=6,7$.

For a representation $\pi$ of $\G(\Phi, R)$ on $V$ we denote by $\Lambda(\pi)$ the set of weights of $\pi$ and by $V_\lambda$ the weight subspace corresponding to $\lambda\in\Lambda(\pi)$.
For any of the representations described above (with the sole exception of $(\rE_8, \varpi_8)$) all weight subspaces $V_\lambda$ are one-dimensional. 
This allows us to choose some basis $E_\pi=\{e_\lambda\}_{\lambda\in\Lambda(\pi)}$.

It will be convenient to use standard numbering for the weights of natural representations of classical groups (cp.~\cite[§1B]{St78}):
$$\begin{array}{cll}
  1,2,\ldots, l+1 & \text{in the case} & \Phi =\rA_l, \\
  1,2,\ldots l, 0, -\ell,\ldots, -2, -1 & \text{in the case} & \Phi =\rB_\ell, \\
  1,2,\ldots l, -\ell,\ldots, -2, -1 & \text{in the cases}   & \Phi =\rC_\ell, \rD_\ell. \\
\end{array}$$
For example, we write $1$ instead of $\varpi_1$, $2$ instead of $\varpi_1-\alpha_1$ and so on.

For a weight $\lambda\in \Lambda(\pi)$ we denote by $\mu_{\lambda}\colon V\rightarrow R$ the coordinate function corresponding $\lambda$.
In other words, $\mu_\lambda(v)$ is the coefficient of the basis vector $e_\lambda$ in the expansion of $v\in V$ in $E_\pi$.
We denote by $v^+$ the basis vector corresponding to highest weight. Using the above notation for classical $\Phi$ we have $v^+=e_1$.
%TODO: State Matsumoto lemma and Chevalley-Matsumoto lemma

@@

\section{Relative parabolic factorizations} \label{sec:factorizations}
\subsection{Relative Bass---Kolster decompositions}\label{sec:bass-kolster}
\input{bass-kolster}
\subsection{Relative Dennis-Vaserstein decompositions}\label{sec:dennis-vaserstein}
\input{dennis-vaserstein}

### 2015-05-06 stability-conditions.tex
Fixed statement of Bass-Kolster; improved treatment of representations

In this section we define stability conditions used in the statements of decomposition results from section~\ref{sec:factorizations}.
First, we recall the notion of stable rank of a ring without unit introduced by L.~Vaserstein in~\cite{Va69, Va71}.
As we are mainly concerned with Chevalley groups, our exposition of stable ranks is neccessarily limited to the case of a commutative ring.
Our next goal is to define the relative version of the absolute stable rank condition which generalizes the definition of~\cite{EO, MKV}.
In the end of this section we state several technical lemmas describing the action of certain unipotent radicals on unimodular columns under stability conditions.

>>>>

In this section we define stability conditions used in the statements of our decomposition theorems from section~\ref{sec:factorizations}.
First, we recall the notion of the stable rank of an ideal introduced by L.~Vaserstein in~\cite{Va69, Va71}.
As we are mainly concerned with applications to Chevalley groups, our exposition of stable ranks is neccessarily limited to commutative rings.
Our next goal is to define the relative version of the absolute stable rank condition which generalizes the definition of~\cite{EO, MKV}.
In the end of this section we state several technical lemmas describing the action of certain unipotent radicals on unimodular columns under stability conditions.
